\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-02-08 doc/publications/plos16/plos16.tex
Many general changes and new reference

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
pronounced in the health space where, for example, whole genome sequencing (WGS) technology enables researchers to
interrogate all 3 billion base pairs of the human genome.

Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs,
termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
separately~\cite{CCC2007}.  It has since been demonstrated that not taking interactions between SNPs into account
when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011}.

A tree based model is capable of taking the interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelize. As the forest is a sum of decision trees, it is possible to grow separate
trees on different processors and combine the trees. See the discussion of the properties of random forests in Section
\ref{section:methods}.

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelization to grow multiple
trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015} written in Go,
achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
multi-threading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
specifically optimized for dealing with large data sets.

The use of traditional compute infrastructure limits the parallelization strategies that can be employed by  
these methods.  The programs are limited to utilizing only CPUs that are on the same computer node (multi-threading) or
else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelization strategies.  In the case of
random forests this means that the computations for each node of a tree can be handed off to separate processors.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest is capable of
parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest outperforms existing tools even on small datasets where multi-threading generally performs
well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
trees and use a high \mtry value which may lead to greater accuracy (see the example in section \ref{section:synthetic}).

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1), the time taken is excessive due to the large amount of data being aggregated and processed by the driver node during
intermediate stages of building the model (see Table~\ref{timingtable}).  This unbalanced work load where the driver
node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. The memory
requirements per executor also increases with dimensionality due to the data types Spark ML uses, which we elaborate on
in the discussion.

\cursedforest is included in our previously developed framework for Spark-based genomic data analysis,
VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be
utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAs catalogue and generate datasets
of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
compute paradigms to deal with these challenges.

In the first section we explore some of the properties of random forests as these will also be apparent in the
\cursedforest implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability
of \cursedforest in respect to the dimensionality of the data, building a random forest model on whole-genome data from
the 1000 Genomes Project. We report the error in predicting ethnicity.  Finally, given the role
different parameter values can play in model construction, we explore the effect tuning these parameters can have on the
prediction accuracy of the model.

@@

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are invariant to monotonic scaling of the data;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to uninformative predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training data sets are
independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
$(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and a tree is grown on each of these $B$ samples.
The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees
over the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
\end{equation*}
In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly
selected subset of the variables.

@@

\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be apparent
in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
sets that are:
\begin{itemize}
\item very wide;
\item all variables have the same number of levels (for example, measurements at different points along the genome);
\item the hypothesized functional relationship is of a form amenable to a decision tree.
\end{itemize}
GWAs data sets are an obvious application areas for this algorithm.

\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interaction of large groups of SNPs.

@@

As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
samples to a dataset, Spark can simply schedule extra tasks to handle the additional items in the RDD.

However, within an RDD, spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
typical datasets, however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
$$
I_G(f) = \sum_{q = 1}^Q f_q ( 1 - f_q ), 
$$
which is a minimum when all observations at the node are in the same class.

@@

\subsubsection{Synthetic data} 
\label{section:synthetic_data}
 Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
probabilities.

The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ and we set 
\begin{equation}
\label{equation:synthetic.data}
z =  \sum_{i=1}^{i=5} {w_i x_i}.
\end{equation}
We let $\sigma_\epsilon^2 = Var(z)(1-\theta)/\theta $  where $\theta$ is a parameter controling the fraction of variance explained by the infomative variables and in our study we chose $\theta = 0.125$ . Then 
$y =   z + \epsilon$ where $\epsilon \sim  N(0, \sigma_\epsilon^2).$
The dichotomous response is generated by thresholding $y$ at the $0.5$ quantile.

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.} In the TCGA example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance for millions of features}
\label{section:synthetic}
In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$) of which 5 
are related to the response variable and 5000 samples ($n=5000$), as discussed in the methods section.

We fit the random forest model and estimate the classification accuracy by capturing the out-of-bag error (OOB) and the
variable selection performance by capturing the rank-biased-overlap (RBO) \cite{Webber2010} measure. RBO compares two
lists (in this case, the paramters $w_i, i=1,\ldots , 5$ (equation \ref{equation:synthetic.data}), and the top 1000
ranked variable importances from the \cursedforest model). It is a mesaure of similarity between rankings that handles
list of different length, weighs high ranks more heavily than low, and is monotonic with increasing depth of
evaluation. We expect to retrieve the 5 features in order of the weight that was assigned to them and with a higher
variable importance measure than the noise variables. RBO provides a quantiative measure of how close we come to this on
a scale from 0 (no feature recovered) to 1 (fully recovered). See supplementary information (S*) for more details.

@@

Increasing \mtry\ in combination with \ntree\ yields best performance with the OOB error essentially constant around 0.4
across a large range of of \mtry and \ntree values.  This is in contrast to the feature-selection performance, where the
RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry.

This may be because a large \mtry\ generates more correlated trees as the same important features have a higher chance
of being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification
tasks where random features can mimic the response variable, hence resulting more in a performance plateau.  
\marginpar{unclear what this means}
Increasing
the number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelising trees at
the node level hence cater perfectly for the requirement of large feature datasets as more trees can be built given a
fixed time budget.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
In this section we use \cursedforest\ to identify genomic locations that are associated with a specific trait. To do
this, we perform feature selection over 7,225,008 genomic variants to identify the locations associated with Bone
Mineral Density (BMD)~\cite{Duncan.2011}. We build a model on the BMD dataset (see datasets) using a binary high or low
BMD value as the label.  From this model, we output a list of the top 10,000 contributing variants, ranked by their Gini
impurity. The variant with the highest Gini impurity, and hence importance, is ranked 1, and the variant with the lowest
Gini impurity is ranked 10,000.

@@

\section{Conclusion}
We have demonstrated that using a different parallelization model can extend random forests to the case of an
extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
the variables are uninformative and have demonstrated the utility of the model for large GWAs data sets. By comparing
this implementation to other implementations (including those optimized for large data sets) we have demonstrated the
utility of this approach.

>>>>

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
life~\cite{Loebbecke2015}.  These datasets are not only growing vertically, by capturing more events, but also
horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
pronounced in the health space where, for example, whole genome sequencing (WGS) technology enables researchers to
interrogate all 3 billion base pairs of the human genome.

Identifying the relevant or disease specific base pair differences between individuals is the focus of genome wide
association studies (GWAS).  These analyses are typically performed on only the most frequently differing base pairs,
termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
separately~\cite{CCC2007}.  It has since been demonstrated that not taking interactions between SNPs into account
when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011}.

A tree-based model is capable of taking the interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelize. As the forest is a sum of decision trees, it is possible to grow separate
trees on different processors and combine the trees. See the discussion of the properties of random forests in Section
\ref{section:methods}.

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelization to grow multiple
trees simultaneously~\cite{Liaw.and.Weiner.2002} and then combine them.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go,
achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
multithreading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
specifically optimized for dealing with large datasets.

The use of traditional compute infrastructure limits the parallelization strategies that can be employed by  
these methods.  The programs are limited to utilizing only CPUs that are on the same computer node (multithreading) or
else they can run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitations by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelization strategies.  In the case of
random forests this means that the computations for each node of a tree can be handed off to separate processors.

Here we introduce \cursedforest, a Hadoop/Spark-based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) datasets. A Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest is capable of
parallelizing the split for each node in a tree thereby handling millions of features. It can, for example, fit a
model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest outperforms existing tools even on small datasets where multithreading generally performs
well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
trees and use a high \mtry value which may lead to greater accuracy (see the example in section \ref{section:synthetic}).

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for a large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1), the time taken is excessive due to the large amount of data being aggregated and processed by the driver node during
intermediate stages of building the model (see Table~\ref{timingtable}).  This unbalanced work load where the driver
node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. The memory
requirements per executor also increases with dimensionality due to the data types Spark ML uses, which we elaborate on
in the discussion.

\cursedforest is included in our previously developed framework for Spark-based genomic data analysis,
VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
comprehensive analysis toolkit that can scale for future data demands, where the richness from WGS data can be
utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAS catalogue and generate datasets
of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
compute paradigms to deal with these challenges.

In the first section we explore some of the properties of random forests as these will also be apparent in the
\cursedforest implementation. Here we provide demonstrations using synthetic data. Secondly, we demonstrate the ability of
\cursedforest to select previously demonstrated features from a bone mineral density (BMD) dataset. 
Thirdly, we demonstrate the scalability of \cursedforest in respect to the dimensionality of the data by building a random
forest model on whole-genome data from the 1000 Genomes Project. Here we report the error in predicting ethnicity.
Finally, given the role different parameter values can play in model construction, we explore the effect that tuning these
parameters can have on the prediction accuracy of the model.

@@

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are invariant to monotonic scaling of the data;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to uninformative predictors;
  \item are able to capture interactions between features (which is important for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training datasets are
independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
$(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and a tree is grown on each of these $B$ samples.
The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees
over the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
\end{equation*}
In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly
selected subset of the variables.

@@

\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be apparent
in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
sets that are:
\begin{itemize}
\item very wide;
\item all variables have the same number of levels (for example, measurements at different points along the genome);
\item the hypothesized functional relationship is of a form amenable to a decision tree.
\end{itemize}
GWAS datasets are an obvious application area for this algorithm.

\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the dataset. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAS application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interactions of large groups of SNPs.

@@

As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
the scalability of Spark as they can be distributed across multiple nodes and operated on in parallel. Even as we add more
samples to a dataset, Spark can simply schedule extra tasks to handle the additional items in the RDD.

However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
typical datasets, however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labelled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
$$
I_G(f) = \sum_{q = 1}^Q f_q ( 1 - f_q ), 
$$
which is at a minimum when all observations at the node are in the same class.

@@

\subsubsection{Synthetic data} 
\label{section:synthetic_data}
 Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomics variation) randomly generated from a uniform distribution with equal
probabilities.

The model parameters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ and we set 
\begin{equation}
\label{equation:synthetic.data}
z =  \sum_{i=1}^{i=5} {w_i x_i}.
\end{equation}
We let $\sigma_\epsilon^2 = Var(z)(1-\theta)/\theta $  where $\theta$ is a parameter controlling the fraction of variance explained by the informative variables and in our study we chose $\theta = 0.125$ . Then 
$y =   z + \epsilon$ where $\epsilon \sim  N(0, \sigma_\epsilon^2).$
The dichotomous response is generated by thresholding $y$ at the $0.5$ quantile.

@@

\subsection{Parameter settings}
We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but give a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance for millions of features}
\label{section:synthetic}
In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$), of which 5 
are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.

We fit the random forest model and estimate the classification accuracy by capturing the out-of-bag error (OOB) and the
variable selection performance by capturing the rank-biased overlap (RBO) \cite{Webber2010} measure. RBO compares two
lists (in this case, the parameters $w_i, i=1,\ldots , 5$ (Equation~\ref{equation:synthetic.data}), and the top 1000
ranked variable importances from the \cursedforest model). It is a mesaure of similarity between rankings that handles
lists of different lengths, weighs high ranks more heavily than low, and is monotonic with increasing depth of
evaluation. We expect to retrieve the 5 features in order of the weight that was assigned to them and with a higher
variable importance measure than the noise variables. RBO provides a quantiative measure of how close we come to this on
a scale from 0 (no feature recovered) to 1 (fully recovered). See supplementary information (S*) for more details.

@@

Increasing \mtry\ in combination with \ntree\ yields the best performance with the OOB error essentially constant around 0.4
across a large range of of \mtry and \ntree values.  This is in contrast to the feature-selection performance, where the
RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry.

This may be because a large \mtry\ generates more correlated trees as the same important features have a higher chance
of being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification
tasks where random features can mimic the response variable, hence resulting more in a performance plateau.  
\marginpar{unclear what this means}
Increasing
the number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelising trees at
the node level hence caters perfectly to the requirement of large feature datasets as more trees can be built given a
fixed time budget.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
In this section we use \cursedforest\ to identify genomic locations that are associated with a specific trait. To do
this, we perform feature selection over 7.2 million genomic variants to identify the locations associated with Bone
Mineral Density (BMD)~\cite{Duncan.2011}. We build a model on the BMD dataset (see datasets) using a binary high or low
BMD value as the label.  From this model, we output a list of the top 10,000 contributing variants, ranked by their Gini
impurity. The variant with the highest Gini impurity, and hence importance, is ranked 1, and the variant with the lowest
Gini impurity is ranked 10,000.

@@

\section{Conclusion}
We have demonstrated that using a different parallelization model can extend random forests to the case of an
extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
the variables are uninformative and have demonstrated the utility of the model for large GWAS datasets. By comparing
this implementation to other implementations (including those optimized for large datasets) we have demonstrated the
utility of this approach.
